---
title: "GAMM and GLMM build"
author: "Silpa Soni Nallacheruvu"
date: "2026-04-01"
output: html_document
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(echo = TRUE)
#pacman::p_load(mgcv, dplyr, readr, lme4)
library(readr)
library(mgcv)
library(lme4)
library(dplyr)
```

## Load the simulated Data

```{r}
catss <- read_csv("catss_final_data.csv")
```

## Check complete cases 

```{r}
print(paste("Number of complete cases:", sum(complete.cases(catss))))
```

## Create the CCA dataset

```{r}
catss <- catss[complete.cases(catss), ]
```

## Split the data into training and testing dataset 

Based on their twin_id

```{r}

test_size <- 0.25

# Get unique IDs
unique_ids <- unique(catss$cmpair)

# Sample test IDs
n_test <- floor(length(unique_ids) * test_size)
test_ids <- sample(unique_ids, size = n_test)

# Split dataset
test_df <- catss[catss$cmpair %in% test_ids, ]
train_df <- catss[!catss$cmpair %in% test_ids, ]

# Print summary
cat(sprintf(
  "Train: %d rows (%d IDs), Test: %d rows (%d IDs)\n",
  nrow(train_df), length(unique(train_df$cmpair)),
  nrow(test_df), length(unique(test_df$cmpair))
))

```


## Prepare the data for the model formula by creating interaction terms and factor variables

```{r}
# get a list of columns with their prefix
SUD15_vars <- grep("^SUD15", names(catss), value = TRUE)
SCZ15_vars <- grep("^SCZ15", names(catss), value = TRUE)
PRS_vars <- grep("^PRS", names(catss), value = TRUE)
ADHD9_vars <- grep("^ADHD9", names(catss), value = TRUE)
ASD9_vars <- grep("^ASD9", names(catss), value = TRUE)
ACE_vars <- grep("^ACE", names(catss), value = TRUE)
SES_vars <- grep("^SES", names(catss), value = TRUE)
PCA_vars <- grep("^PC", names(catss), value = TRUE)[1:2]
Sex_var <- grep("^SEX", names(catss), value = TRUE)
Batch_var <- grep("^batch", names(catss), value = TRUE)

factor_terms <- paste(
  c(
    paste0("factor(", SUD15_vars, ")")
  ),
  collapse = " + "
)


covariate_terms <- paste(
  c(SUD15_vars, SCZ15_vars, ADHD9_vars, ASD9_vars, ACE_vars, SES_vars, PRS_vars, Sex_var),
  collapse = " + "
)

# PCA terms according to the genetic batch
genetic_confounding_terms <- paste(
  c(
    paste0("factor(", Batch_var, ") : ", PCA_vars)
  ),
  collapse = " + "
)

# create interaction terms between SUD15 variables and all other covariates
factor_interaction_terms <- paste(
  c(
    paste0("factor(", SUD15_vars, "):(", SCZ15_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", ADHD9_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", ASD9_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", ACE_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", SES_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", PRS_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", Sex_var, ")")
  ),
  collapse = " + "
)

interaction_terms <- paste(
  c(
    paste0(SUD15_vars ,":", SCZ15_vars),
    paste0(SUD15_vars, ":", ADHD9_vars),
    paste0(SUD15_vars, ":", ASD9_vars),
    paste0(SUD15_vars, ":", ACE_vars),
    paste0(SUD15_vars, ":", SES_vars),
    paste0(SUD15_vars, ":", PRS_vars),
    paste0(SUD15_vars, ":", Sex_var)
  ),
  collapse = " + "
)

train_df$cmpair <- as.factor(train_df$cmpair)
```

# Fit the GAMM model

## Fit the Positive GAMM model

```{r}
pos_gamm_formula_str <- paste(
  "SCZ18_Pos_Norm ~",
  covariate_terms, "+",
  genetic_confounding_terms, "+",
  interaction_terms, "+",
  "s(cmpair, bs='re')"
)

pos_gamm_formula <- as.formula(pos_gamm_formula_str)

pos_gamm_model <- gam(
  pos_gamm_formula,
  data = train_df,
  family = gaussian(),
  method = "REML"
)

summary(pos_gamm_model)
```

## Check the twin random effect

```{r}
plot(pos_gamm_model, select = 1, main = "Random Effect of Twin ID in Positive GAMM Model")
```

# Evaluate the model predictions

```{r}
pos_gamm_y_pred <- predict(pos_gamm_model, newdata = test_df, type = "response")
pos_y_true <- test_df$SCZ18_Pos_Norm


pos_gamm_rmse <- sqrt(mean((pos_y_true - pos_gamm_y_pred)^2))
pos_gamm_mae <- mean(abs(pos_y_true - pos_gamm_y_pred))
pos_gamm_r2 <- 1 - sum((pos_y_true - pos_gamm_y_pred)^2) / sum((pos_y_true - mean(pos_y_true))^2)
pos_gamm_corr <- cor(pos_y_true, pos_gamm_y_pred)

cat("Positive GAMM RMSE:", pos_gamm_rmse, "\n")
cat("Positive GAMM MAE:", pos_gamm_mae, "\n")
cat("Positive GAMM R2:", pos_gamm_r2, "\n")
cat("Positive GAMM Correlation:", pos_gamm_corr, "\n")
```

# Plot predicted vs true values

```{r}
plot(pos_y_true, pos_gamm_y_pred, main = "Predicted vs True Values in Positive GAMM Model", xlab = "True Values", ylab = "Predicted Values")
abline(a = 0, b = 1, col = "red") # indicating perfect predictions
```


## Fit the Negative GAMM model

```{r}
neg_gamm_formula_str <- paste(
  "SCZ18_Neg_Norm ~",
  covariate_terms, "+",
  genetic_confounding_terms, "+",
  interaction_terms, "+",
  "s(cmpair, bs='re')"
)

neg_gamm_formula <- as.formula(neg_gamm_formula_str)

neg_gamm_model <- gam(
  neg_gamm_formula,
  data = train_df,
  family = gaussian(),
  method = "REML"
)

summary(neg_gamm_model)
```

## Check the twin random effect

```{r}
plot(neg_gamm_model, select = 1, main = "Random Effect of Twin ID in Negative GAMM Model")
```

# Evaluate the model predictions

```{r}
neg_gamm_y_pred <- predict(neg_gamm_model, newdata = test_df, type = "response")
neg_y_true <- test_df$SCZ18_Neg_Norm


neg_gamm_rmse <- sqrt(mean((neg_y_true - neg_gamm_y_pred)^2))
neg_gamm_mae <- mean(abs(neg_y_true - neg_gamm_y_pred))
neg_gamm_r2 <- 1 - sum((neg_y_true - neg_gamm_y_pred)^2) / sum((neg_y_true - mean(neg_y_true))^2)
neg_gamm_corr <- cor(neg_y_true, neg_gamm_y_pred)

cat("Negative GAMM RMSE:", neg_gamm_rmse, "\n")
cat("Negative GAMM MAE:", neg_gamm_mae, "\n")
cat("Negative GAMM R2:", neg_gamm_r2, "\n")
cat("Negative GAMM Correlation:", neg_gamm_corr, "\n")
```

# Plot predicted vs true values

```{r}
plot(neg_y_true, neg_gamm_y_pred, main = "Predicted vs True Values in Negative GAMM Model", xlab = "True Values", ylab = "Predicted Values")
abline(a = 0, b = 1, col = "red") # indicating perfect predictions
```


# Fit a GLMM Model

## Fit a Positive GLMM Model

```{r}
pos_glmm_formula_str <- paste(
  "SCZ18_Pos_Norm ~",
  factor_terms, "+",
  covariate_terms, "+",
  interaction_terms, "+",
  "(1 | cmpair)"
)

pos_glmm_formula <- as.formula(pos_glmm_formula_str)

pos_glmm_model <- glmer(
  pos_glmm_formula,
  data = train_df
)

summary(pos_glmm_model)
```

## Check the twin random effect

```{r}
plot(pos_glmm_model, select = 1, main = "Random Effect of Twin ID in Positive GLMM Model")
```

## Evaluate the model predictions

```{r}
pos_glmm_y_pred <- predict(pos_glmm_model, newdata = test_df, allow.new.levels = TRUE, type = "response")
pos_y_true <- test_df$SCZ18_Pos_Norm


pos_glmm_rmse <- sqrt(mean((pos_y_true - pos_glmm_y_pred)^2))
pos_glmm_mae <- mean(abs(pos_y_true - pos_glmm_y_pred))
pos_glmm_r2 <- 1 - sum((pos_y_true - pos_glmm_y_pred)^2) / sum((pos_y_true - mean(pos_y_true))^2)
pos_glmm_corr <- cor(pos_y_true, pos_glmm_y_pred)

cat("Positive GLMM RMSE:", pos_glmm_rmse, "\n")
cat("Positive GLMM MAE:", pos_glmm_mae, "\n")
cat("Positive GLMM R2:", pos_glmm_r2, "\n")
cat("Positive GLMM Correlation:", pos_glmm_corr, "\n")
```

## Plot predicted vs true values

```{r}
plot(pos_y_true, pos_glmm_y_pred, main = "Predicted vs True Values in Positive GLMM Model", xlab = "True Values", ylab = "Predicted Values")
abline(a = 0, b = 1, col = "red") # indicating perfect predictions
```


## Fit a Negative GLMM Model

```{r}
neg_glmm_formula_str <- paste(
  "SCZ18_Neg_Norm ~",
  factor_terms, "+",
  covariate_terms, "+",
  interaction_terms, "+",
  "(1 | cmpair)"
)

neg_glmm_formula <- as.formula(neg_glmm_formula_str)

neg_glmm_model <- glmer(
  neg_glmm_formula,
  data = train_df
)

summary(neg_glmm_model)
```

## Check the twin random effect

```{r}
plot(neg_glmm_model, select = 1, main = "Random Effect of Twin ID in Negative GLMM Model")
```

## Evaluate the model predictions

```{r}
neg_glmm_y_pred <- predict(pos_glmm_model, newdata = test_df, allow.new.levels = TRUE, type = "response")
neg_y_true <- test_df$SCZ18_Pos_Norm


neg_glmm_rmse <- sqrt(mean((neg_y_true - neg_glmm_y_pred)^2))
neg_glmm_mae <- mean(abs(neg_y_true - neg_glmm_y_pred))
neg_glmm_r2 <- 1 - sum((neg_y_true - neg_glmm_y_pred)^2) / sum((neg_y_true - mean(neg_y_true))^2)
neg_glmm_corr <- cor(neg_y_true, neg_glmm_y_pred)

cat("Negative GLMM RMSE:", neg_glmm_rmse, "\n")
cat("Negative GLMM MAE:", neg_glmm_mae, "\n")
cat("Negative GLMM R2:", neg_glmm_r2, "\n")
cat("Negative GLMM Correlation:", neg_glmm_corr, "\n")
```

## Plot predicted vs true values

```{r}
plot(neg_y_true, neg_glmm_y_pred, main = "Predicted vs True Values in Negative GLMM Model", xlab = "True Values", ylab = "Predicted Values")
abline(a = 0, b = 1, col = "red") # indicating perfect predictions
```

