---
title: "GAMM and GLMM toy build"
author: "Silpa Soni Nallacheruvu"
date: "2026-04-01"
output: html_document
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(mgcv, dplyr, readr)
library(readr)
library(mgcv)
library(lme4)
library(dplyr)
```

## Load the simulated Data

```{r}
catss <- read.csv("../Data/simulated_data.csv")
```

## Check complete cases 

```{r}
print(paste("Number of complete cases:", sum(complete.cases(catss))))
```

## Create the output

```{r}
catss <- catss[complete.cases(catss), ]
catss <- catss %>%
  mutate(SCZ18_sum = rowSums(select(., starts_with("SCZ18")), na.rm = TRUE))
```

## Split the data into training and testing dataset 

Based on their twin_id

```{r}

test_size <- 0.25

# Get unique IDs
unique_ids <- unique(catss$twin_id)

# Sample test IDs
n_test <- floor(length(unique_ids) * test_size)
test_ids <- sample(unique_ids, size = n_test)

# Split dataset
test_df <- catss[catss$twin_id %in% test_ids, ]
train_df <- catss[!catss$twin_id %in% test_ids, ]

# Print summary
cat(sprintf(
  "Train: %d rows (%d IDs), Test: %d rows (%d IDs)\n",
  nrow(train_df), length(unique(train_df$twin_id)),
  nrow(test_df), length(unique(test_df$twin_id))
))

```


## Prepare the data for the model formula by creating interaction terms and factor variables

```{r}
# get a list of columns with their prefix
SUD15_vars <- grep("^SUD15_var_", names(catss), value = TRUE)
SCZ15_vars <- grep("^SCZ15_var_", names(catss), value = TRUE)
PRS_vars <- grep("^PRS_var_", names(catss), value = TRUE)
ADHD9_vars <- grep("^ADHD9_var_", names(catss), value = TRUE)
ASD9_vars <- grep("^ASD9_var_", names(catss), value = TRUE)
ACE_vars <- grep("^ACE_var_", names(catss), value = TRUE)
SES_vars <- grep("^SES_var_", names(catss), value = TRUE)
PCA_vars <- grep("^PCA_var_", names(catss), value = TRUE)
Sex_var <- grep("^Sex_var_", names(catss), value = TRUE)

factor_terms <- paste(
  c(
    paste0("factor(", SUD15_vars, ")")
  ),
  collapse = " + "
)


covariate_terms <- paste(
  c(SCZ15_vars, ADHD9_vars, ASD9_vars, ACE_vars, SES_vars, PCA_vars, PRS_vars, Sex_var),
  collapse = " + "
)

# create interaction terms between SUD15 variables and all other covariates
interaction_terms <- paste(
  c(
    paste0("factor(", SUD15_vars, "):(", SCZ15_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", ADHD9_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", ASD9_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", ACE_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", SES_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", PRS_vars, ")"),
    paste0("factor(", SUD15_vars, "):(", Sex_var, ")")
  ),
  collapse = " + "
)
```

# Fit the GAMM model

```{r}
gamm_formula_str <- paste(
  "SCZ18_sum ~",
  factor_terms, "+",
  covariate_terms, "+",
  interaction_terms, "+",
  "s(twin_id, bs='re')"
)

gamm_formula <- as.formula(gamm_formula_str)

gamm_model <- gam(
  gamm_formula,
  data = train_df,
  family = gaussian(),
  method = "REML"
)

summary(gamm_model)
```

## Check the twin random effect

```{r}
plot(gamm_model, select = 1, main = "Random Effect of Twin ID in GAMM Model")
```

# Evaluate the model predictions

```{r}
gamm_y_pred <- predict(gamm_model, newdata = test_df, type = "response")
y_true <- test_df$SCZ18_sum


gamm_rmse <- sqrt(mean((y_true - gamm_y_pred)^2))
gamm_mae <- mean(abs(y_true - gamm_y_pred))
gamm_r2 <- 1 - sum((y_true - gamm_y_pred)^2) / sum((y_true - mean(y_true))^2)
gamm_corr <- cor(y_true, gamm_y_pred)

cat("GAMM RMSE:", gamm_rmse, "\n")
cat("GAMM MAE:", gamm_mae, "\n")
cat("GAMM R2:", gamm_r2, "\n")
cat("GAMM Correlation:", gamm_corr, "\n")
```
# Plot predicted vs true values

```{r}
plot(y_true, gamm_y_pred, main = "Predicted vs True Values in GAMM Model", xlab = "True Values", ylab = "Predicted Values")
abline(a = 0, b = 1, col = "red") # indicating perfect predictions
```


# Fit a GLMM model

```{r}
glmm_formula_str <- paste(
  "SCZ18_sum ~",
  factor_terms, "+",
  covariate_terms, "+",
  interaction_terms, "+",
  "(1 | twin_id)"
)

glmm_formula <- as.formula(glmm_formula_str)

glmm_model <- glmer(
  glmm_formula,
  data = train_df
)

summary(glmm_model)
```

## Check the twin random effect

```{r}
plot(glmm_model, select = 1, main = "Random Effect of Twin ID in GLMM Model")
```

## Evaluate the model predictions

```{r}
glmm_y_pred <- predict(glmm_model, newdata = test_df, allow.new.levels = TRUE, type = "response")
y_true <- test_df$SCZ18_sum


glmm_rmse <- sqrt(mean((y_true - glmm_y_pred)^2))
glmm_mae <- mean(abs(y_true - glmm_y_pred))
glmm_r2 <- 1 - sum((y_true - glmm_y_pred)^2) / sum((y_true - mean(y_true))^2)
glmm_corr <- cor(y_true, glmm_y_pred)

cat("GLMM RMSE:", glmm_rmse, "\n")
cat("GLMM MAE:", glmm_mae, "\n")
cat("GLMM R2:", glmm_r2, "\n")
cat("GLMM Correlation:", glmm_corr, "\n")
```

## Plot predicted vs true values

```{r}
plot(y_true, glmm_y_pred, main = "Predicted vs True Values in GLMM Model", xlab = "True Values", ylab = "Predicted Values")
abline(a = 0, b = 1, col = "red") # indicating perfect predictions
```

