---
title: "GAMM model for Psychotic Risk prediction from SUbstance Use and other factors in CATSS"
author: "Silpa Soni Nallacheruvu"
date: "2026-04-01"
output: html_document
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(mgcv)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(patchwork)
```

## Load the simulated Data

```{r}
catss <- read_csv("catss_final_data.csv")
```

## Check complete cases 

```{r}
print(paste("Number of complete cases:", sum(complete.cases(catss))))
```

## Create the CCA dataset

```{r}
catss <- catss[complete.cases(catss), ]
```

## Split the data into training and testing dataset 

Based on their twin_id

```{r}

test_size <- 0.25

# Get unique IDs
unique_ids <- unique(catss$cmpair)

# Sample test IDs
n_test <- floor(length(unique_ids) * test_size)
test_ids <- sample(unique_ids, size = n_test)

# Split dataset
test_df <- catss[catss$cmpair %in% test_ids, ]
train_df <- catss[!catss$cmpair %in% test_ids, ]

# Print summary
cat(sprintf(
  "Train: %d rows (%d IDs), Test: %d rows (%d IDs)\n",
  nrow(train_df), length(unique(train_df$cmpair)),
  nrow(test_df), length(unique(test_df$cmpair))
))

```


## Prepare the data for the model formula by creating interaction terms and factor variables

```{r}
# get a list of columns with their prefix
SUD15_vars <- grep("^SUD15", names(catss), value = TRUE)
SCZ15_vars <- grep("^SCZ15", names(catss), value = TRUE)
PRS_vars <- grep("^PRS", names(catss), value = TRUE)
ADHD9_vars <- grep("^ADHD9", names(catss), value = TRUE)
ASD9_vars <- grep("^ASD9", names(catss), value = TRUE)
ACE_vars <- grep("^ACE", names(catss), value = TRUE)
SES_vars <- grep("^SES", names(catss), value = TRUE)
PCA_vars <- grep("^PC", names(catss), value = TRUE)[1:2]
Sex_var <- grep("^SEX", names(catss), value = TRUE)
Batch_var <- grep("^batch", names(catss), value = TRUE)[1]

covariate_terms <- paste(
  c(SUD15_vars, SCZ15_vars, ADHD9_vars, ASD9_vars, ACE_vars, SES_vars, PRS_vars, Sex_var),
  collapse = " + "
)

# PCA terms according to the genetic batch
genetic_confounding_terms <- paste(
  c(
    paste0("factor(", Batch_var, ") : ", PCA_vars)
  ),
  collapse = " + "
)

# create interaction terms between SUD15 variables and all other covariates
interaction_terms <- paste(
  c(
    paste0(SUD15_vars ,":", SCZ15_vars),
    paste0(SUD15_vars, ":", ADHD9_vars),
    paste0(SUD15_vars, ":", ASD9_vars),
    paste0(SUD15_vars, ":", ACE_vars),
    paste0(SUD15_vars, ":", SES_vars),
    paste0(SUD15_vars, ":", PRS_vars),
    paste0(SUD15_vars, ":", Sex_var)
  ),
  collapse = " + "
)

train_df$cmpair <- as.factor(train_df$cmpair)
```

# Fit the GAMM model

## Fit the Positive GAMM model

```{r}
pos_gamm_formula_str <- paste(
  "SCZ18_Pos_Norm ~",
  covariate_terms, "+",
  genetic_confounding_terms, "+",
  interaction_terms, "+",
  "s(cmpair, bs='re')"
)

pos_gamm_formula <- as.formula(pos_gamm_formula_str)

pos_gamm_model <- gam(
  pos_gamm_formula,
  data = train_df,
  family = gaussian(),
  method = "REML"
)

summary(pos_gamm_model)
```


### View significance

```{r}
param_coefs <- as.data.frame(summary(pos_gamm_model)$p.table)

smooth_coefs <- as.data.frame(summary(pos_gamm_model)$s.table)

# Prepare clean table
pos_coefs_table <- param_coefs %>%
  mutate(Variable = rownames(.)) %>%
  rename(p_value = `Pr(>|t|)`) %>%
  mutate(significance = cut(
    p_value,
    breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
    labels = c("***", "**", "*", ".", "")
  )) %>%
  filter(p_value < 0.05, Variable != "(Intercept)") %>%
  arrange(p_value) %>%
  select(Estimate, `Std. Error`, p_value, significance) %>%
  mutate(
    Estimate = round(Estimate, 3),
    `Std. Error` = round(`Std. Error`, 3),
    p_value = signif(p_value, 3)
  )

kable(pos_coefs_table, format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```


### Check the twin random effect

```{r}
plot(pos_gamm_model, select = 1, main = "Random Effect of Twin ID in Positive GAMM Model")
```

### Evaluate the model predictions

```{r}
pos_gamm_y_pred <- predict(pos_gamm_model, newdata = test_df, type = "response")
pos_y_true <- test_df$SCZ18_Pos_Norm

# Create dataframe
pos_gamm_df <- data.frame(
  actual = pos_y_true,
  predicted = pos_gamm_y_pred
)

# Residuals
pos_gamm_df$residuals <- pos_gamm_df$actual - pos_gamm_df$predicted

# Metrics
r2 <- cor(pos_gamm_df$actual, pos_gamm_df$predicted)^2
mae <- mean(abs(pos_gamm_df$actual - pos_gamm_df$predicted))
pearson_r <- cor(pos_gamm_df$actual, pos_gamm_df$predicted)
spearman_r <- cor(pos_gamm_df$actual, pos_gamm_df$predicted, method = "spearman")

metrics_text <- sprintf(
  "R² = %.4f\nMAE = %.4f\nSpearman rho = %.4f\nPearson R = %.4f",
  r2, mae, spearman_r, pearson_r
)

# predicted vs actual plot
p1 <- ggplot(pos_gamm_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.5, color = "#2C7FB8") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  
  annotate("text",
           x = max(pos_gamm_df$actual) * 0.6,
           y = max(pos_gamm_df$predicted) * 0.1,   # move inside plot
           label = metrics_text,
           hjust = 0,
           size = 4) +
  
  coord_cartesian(clip = "off") +  # prevents clipping
  
  labs(
    title = "Positive GAMM Model - Predicted vs Actual (Test set)",
    x = "Actual Values",
    y = "Predicted Values"
  ) +
  theme_minimal() +
  theme(plot.margin = margin(10, 40, 10, 10))  # extra space

p1 

# residual plot
p2 <- ggplot(pos_gamm_df, aes(x = predicted, y = residuals)) +
  geom_point(alpha = 0.5, color = "#74A9CF") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  
  labs(
    title = "Positive GAMM Model - Residual Plot (Test set)",
    x = "Predicted Values",
    y = "Residuals (Actual - Predicted)"
  ) +
  theme_minimal()

p2
```


## Fit the Negative GAMM model

```{r}
neg_gamm_formula_str <- paste(
  "SCZ18_Neg_Norm ~",
  covariate_terms, "+",
  genetic_confounding_terms, "+",
  interaction_terms, "+",
  "s(cmpair, bs='re')"
)

neg_gamm_formula <- as.formula(neg_gamm_formula_str)

neg_gamm_model <- gam(
  neg_gamm_formula,
  data = train_df,
  family = gaussian(),
  method = "REML"
)

summary(neg_gamm_model)
```


### View significance

```{r}
param_coefs <- as.data.frame(summary(neg_gamm_model)$p.table)

smooth_coefs <- as.data.frame(summary(neg_gamm_model)$s.table)

# Prepare clean table
neg_coefs_table <- param_coefs %>%
  mutate(Variable = rownames(.)) %>%
  rename(p_value = `Pr(>|t|)`) %>%
  mutate(significance = cut(
    p_value,
    breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
    labels = c("***", "**", "*", ".", "")
  )) %>%
  filter(p_value < 0.05, Variable != "(Intercept)") %>%
  arrange(p_value) %>%
  select(Estimate, `Std. Error`, p_value, significance) %>%
  mutate(
    Estimate = round(Estimate, 3),
    `Std. Error` = round(`Std. Error`, 3),
    p_value = signif(p_value, 3)
  )

kable(neg_coefs_table, format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```


### Check the twin random effect

```{r}
plot(neg_gamm_model, select = 1, main = "Random Effect of Twin ID in Negative GAMM Model")
```

### Evaluate the model predictions

```{r}
neg_gamm_y_pred <- predict(neg_gamm_model, newdata = test_df, type = "response")
neg_y_true <- test_df$SCZ18_Neg_Norm

# Create data frame
neg_gamm_df <- data.frame(
  actual = neg_y_true,
  predicted = neg_gamm_y_pred
)

# Residuals
neg_gamm_df$residuals <- neg_gamm_df$actual - neg_gamm_df$predicted

# Metrics
r2 <- cor(neg_gamm_df$actual, neg_gamm_df$predicted)^2
mae <- mean(abs(neg_gamm_df$actual - neg_gamm_df$predicted))
pearson_r <- cor(neg_gamm_df$actual, neg_gamm_df$predicted)
spearman_r <- cor(neg_gamm_df$actual, neg_gamm_df$predicted, method = "spearman")

metrics_text <- sprintf(
  "R² = %.4f\nMAE = %.4f\nSpearman rho = %.4f\nPearson R = %.4f",
  r2, mae, spearman_r, pearson_r
)

# predicted vs actual plot
p1 <- ggplot(neg_gamm_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.5, color = "#2C7FB8") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  
  annotate("text",
           x = max(neg_gamm_df$actual) * 0.6,
           y = max(neg_gamm_df$predicted) * 0.1,   # move inside plot
           label = metrics_text,
           hjust = 0,
           size = 4) +
  
  coord_cartesian(clip = "off") +  # prevents clipping
  
  labs(
    title = "Negative GAMM Model - Predicted vs Actual (Test set)",
    x = "Actual Values",
    y = "Predicted Values"
  ) +
  theme_minimal() +
  theme(plot.margin = margin(10, 40, 10, 10))  # extra space

p1

# residual plot
p2 <- ggplot(neg_gamm_df, aes(x = predicted, y = residuals)) +
  geom_point(alpha = 0.5, color = "#74A9CF") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  
  labs(
    title = "Negative GAMM Model - Residual Plot (Test set)",
    x = "Predicted Values",
    y = "Residuals (Actual - Predicted)"
  ) +
  theme_minimal()

p2
```
